The state learner -- a super learner for right-censored data

### 2025-01-13 manuscript/manuscript-lda.tex
manuscript template for lda

\documentclass[11pt]{article}
\usepackage[a4paper,left=2.8cm, right=2.8cm, top=3cm, bottom=3cm]{geometry}

%% Packages
\usepackage{amsthm, dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}

\usepackage[authoryear]{natbib}
\bibliographystyle{abbrvnat}

\definecolor{linkcolor}{rgb}{0, 0, 0.54}
\usepackage[colorlinks,allcolors=linkcolor]{hyperref}

\theoremstyle{thmstyleone}%\
\newtheorem{theorem}{Theorem}%  meant for continuous numbers
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%
\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}

%% Line spacing
\linespread{2}

% \usepackage{lineno}
% \linenumbers

\usepackage{endfloat}

%% Only for comments:
\usepackage[author=]{fixme}
\fxusetheme{color}
\definecolor{fxtarget}{rgb}{.5,.5,.5}
\definecolor{fxnote}{rgb}{.5,.5,.5}
\fxsetup{status=draft}

%% Toggle with or without figures
% \def\nofig{1}

\RequirePackage{graphicx}
\usepackage[font=small]{caption}

\newcommand{\includeFigCond}[2][]{
  \ifx\nofig\undefined %
    \includegraphics[#1]{#2} %
  \else %
    \texttt{#2} %
  \fi %
}

% New operators and commands
\DeclareMathOperator{\E}{\mathbb{E}} % expectation
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}
\renewcommand{\S}{\mathbb{S}}
\newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\renewcommand{\phi}{\varphi}
\renewcommand{\epsilon}{\varepsilon}
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand{\weakly}{\rightsquigarrow}
\newcommand\smallO{\textit{o}}
\newcommand\bigO{\textit{O}}
\newcommand{\midd}{\; \middle|\;}
\newcommand{\1}{\mathds{1}}
\usepackage{ifthen} %% Empirical process with default argument
\newcommand{\G}[2][n]{
{\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}
\newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
\newcommand{\data}{\ensuremath{\mathcal{D}}}

%% Title
\title{\Large The state learner -- a super learner for right-censored data}
\date{\small \today}

\usepackage{authblk}
\author[1,*]{\normalsize Anders Munch}
\author[1]{\normalsize Thomas~A.~Gerds}
\affil[1]{\normalsize Section of Biostatistics, University of Copenhagen}
\affil[*]{\normalsize Address for correspondence: \href{email:email-id.com}{a.munch@sund.ku.dk}}

\usepackage{fancyhdr}
\fancyhf{}
\fancyhead[l]{\textit{Munch et al.}}
\fancyhead[r]{\textit{The state learner}}
\fancyfoot[c]{\thepage}
\renewcommand\headrulewidth{0pt}
\pagestyle{fancy}

\begin{document}

@@

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone tools and in
  applications of causal inference to estimate nuisance parameters. The super
  learner is a machine learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. In right-censored
  data, the choice of the loss function and the estimation of the expected loss
  need careful consideration. We introduce the state learner, a new super
  learner for survival analysis, which simultaneously evaluates libraries of
  prediction models for the event of interest and the censoring distribution.
  The state learner can be applied to all types of survival models, works in the
  presence of competing risks, and does not require a single pre-specified
  estimator of the conditional censoring distribution. We establish an oracle
  inequality for the state learner and investigate its performance through
  numerical experiments. We illustrate the application of the state learner with
  prostate cancer data, as a stand-alone prediction tool, and, for causal
  inference, as a way to estimate the nuisance parameter models of a smooth
  statistical functional.
\end{abstract}

% \keywords{Competing risks, cross-validation, loss based estimation,
% right-censored data, super learner}

\textit{{\small \textbf{Keywords:} Competing risks, cross-validation, loss based
    estimation, right-censored data, super learner}}

\pagebreak

@@

\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
  \caption[]{For the risk prediction models provided by each of the super
    learners, the IPA is plotted against sample size. The results are averages across 
    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
    uncertainty.
  }
  \label{fig:ipcw-fail}
\end{figure}

@@

The results are shown in Figures~\ref{fig:zelefski-out}
and~\ref{fig:zelefski-cens}. We see that for most sample sizes, the state
learner selected prediction models for both censoring and outcome which have
similar or higher IPA compared to the prediction models selected by
\texttt{survSL}.
\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL-out.pdf}
  \caption[]{For the risk prediction models of the outcome provided by each
    of the super learners, the IPA at the fixed time horizon is plotted against
    sample size. The results are averages across 1000 repetitions and the error
    bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski-out}
\end{figure}

\begin{figure}
  \centering %
  \includeFigCond[width=.7\linewidth]{experiment-fig-sl-survSL-cens.pdf}
  \caption[]{For the risk prediction models of the censoring model provided by
    each of the super learners, the IPA at the fixed time horizon is plotted
    against sample size. The results are averages across 1000 repetitions and
    the error bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski-cens}
\end{figure}

@@

\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{real-data-state-learner.pdf}
  \caption[]{The results of applying the 125 combinations of learners to the
    prostate cancer data set. The error bars are based on five repetitions using
    different splits. We refer to learners of \( \Lambda_1 \), \( \Lambda_2 \),
    and $\Gamma$ as `Tumor learner', `Mortality learner', and `Censoring
    learner', respectively.}
  \label{fig:zelefski-real}
\end{figure}

@@

\begin{figure}
  \centering%
  \includeFigCond[width=1\linewidth]{real-data-target.pdf}
  \caption[]{Estimates of the average treatment effect of hormone therapy on the
    risk of tumor recurrence and death obtained from the prostate cancer study
    analysed by \cite{kattan2000pretreatment}. The estimates are based on the
    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with pointwise
    95\% confidence intervals calculated as described in
    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
    parameters are provided by the state learner.}
  \label{fig:zelefski-real-target}
\end{figure}

@@

\section{Competing interests}
No competing interest is declared.

\bibliography{bib.bib}

@@

\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{dsfont, amssymb,soul,xcolor, enumitem, amsmath,verbatim}
\usepackage[authoryear]{natbib}
\definecolor{linkcolor}{rgb}{0, 0, 0.54}
\usepackage[colorlinks,allcolors=linkcolor]{hyperref}
%
% please place your own definitions here and don't use \def but
\DeclareMathOperator{\E}{\mathbb{E}} % expectation
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\C}{\mathbb{C}}
\renewcommand{\S}{\mathbb{S}}
\newcommand{\blank}{\makebox[1ex]{\textbf{$\cdot$}}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\renewcommand{\phi}{\varphi}
\renewcommand{\epsilon}{\varepsilon}
\newcommand*\diff{\mathop{}\!\mathrm{d}}
\newcommand{\weakly}{\rightsquigarrow}
\newcommand\smallO{\textit{o}}
\newcommand\bigO{\textit{O}}
\newcommand{\midd}{\; \middle|\;}
\newcommand{\1}{\mathds{1}}
\usepackage{ifthen} %% Empirical process with default argument
\newcommand{\G}[2][n]{
{\ensuremath{\mathbb{G}_{#1}}{\left[#2\right]}}
}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}
\newcommand{\empmeas}{\ensuremath{\mathbb{P}_n}} % empirical measure
\newcommand{\data}{\ensuremath{\mathcal{D}}}
%
% Insert the name of "your journal" with
\journalname{Lifetime Data Analysis}
%
\begin{document}

\title{The state learner -- a super learner for right-censored data
  % \thanks{Grants or other notes about the article that should go on the front
  % page should be placed here. General acknowledgments should be placed at the
  % end of the article.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}

\titlerunning{The state learner}        % if too long for running head

\author{Anders Munch         \and
        Thomas~A.~Gerds %etc.
}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Anders Munch \at
              Section of Biostatistics, University of Copenhagen \\
              % Tel.: +123-45-678910\\
              % Fax: +123-45-678910\\
              \email{a.munch@sund.ku.dk}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
              \and %
              Thomas~A.~Gerds %
              \at Section of Biostatistics, University of Copenhagen }

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

@@

\begin{abstract}
  In survival analysis, prediction models are needed as stand-alone tools and in
  applications of causal inference to estimate nuisance parameters. The super
  learner is a machine learning algorithm which combines a library of prediction
  models into a meta learner based on cross-validated loss. In right-censored
  data, the choice of the loss function and the estimation of the expected loss
  need careful consideration. We introduce the state learner, a new super
  learner for survival analysis, which simultaneously evaluates libraries of
  prediction models for the event of interest and the censoring distribution.
  The state learner can be applied to all types of survival models, works in the
  presence of competing risks, and does not require a single pre-specified
  estimator of the conditional censoring distribution. We establish an oracle
  inequality for the state learner and investigate its performance through
  numerical experiments. We illustrate the application of the state learner with
  prostate cancer data, as a stand-alone prediction tool, and, for causal
  inference, as a way to estimate the nuisance parameter models of a smooth
  statistical functional.

  \keywords{Competing risks, cross-validation, loss based estimation,
    right-censored data, super learner}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

\begin{figure}
  \centering %
  \includegraphics[width=1\linewidth]{experiment-fig-sl-ipcw.pdf}
  \caption[]{For the risk prediction models provided by each of the super
    learners, the IPA is plotted against sample size. The results are averages across 
    1000 simulated data sets and the error bars are used to quantify the Monte Carlo
    uncertainty.
  }
  \label{fig:ipcw-fail}
\end{figure}

@@

The results are shown in Figures~\ref{fig:zelefski-out}
and~\ref{fig:zelefski-cens}. We see that for most sample sizes, the state
learner selected prediction models for both censoring and outcome which have
similar or higher IPA compared to the prediction models selected by
\texttt{survSL}.
\begin{figure}
  \centering %
  \includegraphics[width=1\linewidth]{experiment-fig-sl-survSL-out.pdf}
  \caption[]{For the risk prediction models of the outcome provided by each
    of the super learners, the IPA at the fixed time horizon is plotted against
    sample size. The results are averages across 1000 repetitions and the error
    bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski-out}
\end{figure}

\begin{figure}
  \centering %
  \includegraphics[width=.7\linewidth]{experiment-fig-sl-survSL-cens.pdf}
  \caption[]{For the risk prediction models of the censoring model provided by
    each of the super learners, the IPA at the fixed time horizon is plotted
    against sample size. The results are averages across 1000 repetitions and
    the error bars are used to quantify the Monte Carlo uncertainty.}
  \label{fig:zelefski-cens}
\end{figure}

@@

\begin{figure}
  \centering %
  \includegraphics[width=1\linewidth]{real-data-state-learner.pdf}
  \caption[]{The results of applying the 125 combinations of learners to the
    prostate cancer data set. The error bars are based on five repetitions using
    different splits. We refer to learners of \( \Lambda_1 \), \( \Lambda_2 \),
    and $\Gamma$ as `Tumor learner', `Mortality learner', and `Censoring
    learner', respectively.}
  \label{fig:zelefski-real}
\end{figure}

@@

\begin{figure}
  \centering%
  \includegraphics[width=1\linewidth]{real-data-target.pdf}
  \caption[]{Estimates of the average treatment effect of hormone therapy on the
    risk of tumor recurrence and death obtained from the prostate cancer study
    analysed by \cite{kattan2000pretreatment}. The estimates are based on the
    estimator defined in equation~(\ref{eq:one-step-comp-ate}) with pointwise
    95\% confidence intervals calculated as described in
    Section~\ref{sec:targeted-learning}. The estimates of the nuisance
    parameters are provided by the state learner.}
  \label{fig:zelefski-real-target}
\end{figure}

@@

\section*{Conflict of interest}

The authors declare that they have no conflict of interest.

% BibTeX users please use one of
% \bibliographystyle{spbasic}      % basic style, author-year citations
%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics

\bibliographystyle{abbrvnat}
\bibliography{bib.bib}   % name your BibTeX data base

@@

\end{document}
% end of file template.tex
